---
title: "DTD Project Final"
author: "Jarryd Wannenburg"
date: "5/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Libraries
```{r include=FALSE}
rm(list = ls())
options(readr.show_col_types = FALSE)
# Set working directory
setwd("C:/Users/jarry/Desktop/Google Data Analytics Cert/bike_share")
# Load packages
library(tidyverse)
library(lubridate)
library(hms)
library(ggplot2)
```


# Loading
Create the raw data frames
```{r eval=FALSE, include=FALSE}
# Create a list of csv files which represent data to be combined after converting to object in excel
files <- list.files(path = 'raw_data/', pattern = ".csv$")
raw <- read_csv(paste0('raw_data/', files[2]))
for (csv in files[-2]) {
  frame <- read_csv(paste0('raw_data/', csv))
  raw <- bind_rows(raw, frame)
}

rm(frame, csv, files)
write_csv(raw, 'data/raw.csv')
```


# Cleaning
## Filtering
manipulate the raw master data frame
```{r include=FALSE}
# Consider coming back and calculating values
if (exists('raw')) {
  if (is.data.frame(get('raw'))) {
    raw_master <- raw %>%
      mutate(
        day_of_week = noquote(weekdays(started_at)),
        trip_length = as_hms(ended_at - started_at),
        trip_length_sec = as.numeric(trip_length) # Easier to deal with seconds than hms format
      ) %>%
      filter(trip_length_sec >= 60, # Remove 61500 trips likely to be test runs or redocking errors (https://ride.divvybikes.com/system-data)
             trip_length_sec < 86400) # Any trip over 24 hours should automatically be removed
    
    rm(raw)
  } else {
    raw_master <- read_csv('data/raw.csv') %>%
      mutate(
        day_of_week = noquote(weekdays(started_at)),
        trip_length = as_hms(ended_at - started_at),
        trip_length_sec = as.numeric(trip_length) # Easier to deal with seconds than hms format
      ) %>%
      filter(trip_length_sec >= 60, # Remove 61500 trips likely to be test runs or redocking errors (https://ride.divvybikes.com/system-data)
             trip_length_sec < 86400) # Any trip over 24 hours should automatically be removed
    
  }
} else{
  raw_master <- read_csv('data/raw.csv') %>%
    mutate(
      day_of_week = noquote(weekdays(started_at)),
      trip_length = as_hms(ended_at - started_at),
      trip_length_sec = as.numeric(trip_length) # Easier to deal with seconds than hms format
    ) %>%
    filter(trip_length_sec >= 60, # Remove 61500 trips likely to be test runs or redocking errors (https://ride.divvybikes.com/system-data)
             trip_length_sec < 86400) # Any trip over 24 hours should automatically be removed
}

# Show the distribution of null values
colSums(is.na(raw_master))

# Bikes don't necessarily need to be picked up or dropped off at a station. But we do need to map the data, so we need a lat and lon
raw_master_no_null <- raw_master[!(is.na(raw_master$end_lat) | raw_master$end_lat=="" | is.na(raw_master$start_lat) | raw_master$start_lat==""), ]
```

```{r eval=FALSE, include=FALSE}
# Write this intermediary df to a file to prevent having to run the data loading and manipulation each time
write_csv(raw_master_no_null,
          'data/dtd_raw_master_wo_null_w_outlier.csv')
write_csv(raw_master, 'data/dtd_raw_master.csv')
```


## Outlier Manipulation
For this analysis, since the data is not normally distributed, outliers will be determined with logic for simplicity, though further analysis is encouraged. Outliers will be deemed as trips likely to be errors as opposed to variation in the data.
```{r include=FALSE}
# Removing outliers once trips with null lat,lon are removed
if (exists('raw_master_no_null')) {
  if (is.data.frame(get('raw_master_no_null'))) {
    print('raw_master_no_null already in environment')
  } else{
    raw_master_no_null <-
      read_csv('data/dtd_raw_master_wo_null_w_outlier.csv')
  }
} else {
  raw_master_no_null <-
    read_csv('data/dtd_raw_master_wo_null_w_outlier.csv')
}

threshold <- 8*60*60 # 8 hours which represents a "full" day trip. With this business models, bikes aren't meant to be kept for long periods.

dtd <- raw_master_no_null %>%
  filter(trip_length_sec <= threshold)
```

```{r eval=FALSE, include=FALSE}
write_csv(dtd, 'data/dtd.csv')
```


# Analysis
## Summary by user type
```{r}
if (exists('dtd')) {
  if (is.data.frame(get('dtd'))) {
    print('dtd already in environment')
  } else{
    dtd <-
      read_csv('data/dtd.csv')
  }
} else {
  dtd <-
    read_csv('data/dtd.csv')
}

# Compare the means
dtd_member <- dtd %>%
  filter(member_casual == 'member')

dtd_casual <- dtd %>%
  filter(member_casual == 'casual')

a <- c("Min.", "Q1", "Median", "Mean", "Q3", "Max")
b <- hms(summary(dtd$trip_length_sec)) %>% unlist() %>% as.numeric() %>% round(0)
c <- hms(summary(dtd_member$trip_length_sec)) %>% unlist() %>% as.numeric() %>% round(0)
d <- hms(summary(dtd_casual$trip_length_sec)) %>% unlist() %>% as.numeric() %>% round(0)

dtd_summary_member_casual_sec <- as.data.frame(cbind(a,b,c,d))
colnames(dtd_summary_member_casual_sec) <- c("Statistic", "Combined", "Member", "Casual")

e <- b %>% hms()
f <- c %>% hms() 
g <- d %>% hms()

dtd_summary_member_casual_hms <- data.frame(a,e,f,g)
colnames(dtd_summary_member_casual_hms) <- c("Statistic", "Combined", "Member", "Casual")

dtd_summary_member_casual_sec
dtd_summary_member_casual_hms

```

```{r}
colSums(is.na(dtd))
write_csv(dtd %>% drop_na(), 'data/dtd_no_null.csv')
```

